Why does preregistration increase the persuasiveness of evidence? A Bayesian rationalization

### 2023-10-12 manuscript.Rmd
small changes after reading

We are interested in why preregistrations should change researchers' evaluation of evidence.
Applying a Bayesian framework allows us to investigate our research question most straightforwardly. 
Specifically, it allows us to model changes in subjective degrees of belief due to preregistration or, more simply, "persuasion".
Please note that our decision to adopt a Bayesian philosophy of science does not make assumptions about the statistical methods researchers use.
In fact, this conceptualization is intentionally as minimal as possible to be compatible with a wide range of philosophies of science and statistical methods researchers might subscribe to.
However, we should note that Popperians would be appalled that we are content with positive inductive inferences (but we dislike "failing to disprove" too much to give it up), and Neopopperians would flinch that we assign probabilities to beliefs (we are fond of calculating things).

@@

The posterior probability $P(H|E)$ is of great relevance since it is often used directly or indirectly as a measure of confirmation of a hypothesis.
In the tradition of Carnap, in its direct use, it is called confirmation as firmness; in its relation to the a priori probability $P(H)$, it is called _increase in firmness_ [@carnapLogicalFoundationsProbability1950, preface to the 1962 edition].
We concentrate on the posterior probability because of its simplicity but take it only as one example of a possible measure.
In reality, surely researchers differ in what function they apply to judge evidence.
The goal is therefore to reason about the space of possible measures researchers might apply.
However, since it is reasonable that any measure of confirmation increases monotonically with an increase in posterior probability $P(H|E)$, we might well take it to illustrate our reasoning.

In short, we want to increase posterior probability $P(H|E)$.
Increases in posterior probability $P(H|E)$ are associated with increased epistemic value, of which we want to maximize the expectation.
So how can we increase posterior probability?
The Bayes formula yields three components that influence confirmation, namely $P(H)$, $P(E|H)$ and $P(E)$.
The first option leads us to the unsurprising conclusion that higher a priori probability $P(H)$ leads to higher posterior probability $P(H|E)$.
If a hypothesis is more probable to begin with, observing evidence in its favor will result in a hypothesis that is more strongly confirmed, all else being equal.
However, the prior probability of a hypothesis is nothing our study design can change.
The second option is equally reasonable; that is, an increase in $P(E|H)$ leads to a higher posterior probability $P(H|E)$.
$P(E|H)$ is the probability of obtaining evidence for a hypothesis when it holds.
We call this probability of detecting evidence, given that the hypothesis holds "detectability."
Consequently, researchers should ensure that their study design allows them to find evidence for their hypothesis, in case it is true.
When applied strictly within the bounds of null hypothesis testing, detectability is equivalent to power (or the complement of type-II error rate).
However, while detectability is of great importance for study design, it is not directly relevant to the objective of preregistration.
However, we latter discuss why detectability can not be disregarded from consideration.
Thus, $P(E)$ remains to be considered.
Since $P(E)$ is the denominator, decreasing it can increase the posterior probability.
In other words, high risk, high reward.

@@

The last statement sounds like a truism but is directly related to Popper's seminal criterion of demarcation.
He stated that if it is impossible to prove that a hypothesis is false ($P(\neg{}E|\neg{}H) = 0$, theoretical risk is zero), it cannot be considered a scientific hypothesis [@popperLogicScientificDiscovery2002, p. 18].
We note these relations to underline that the Bayesian rationale we apply here is able to reconstruct many commonly held views on riskiness and epistemic value (but Popper would disagree with attaching any form of believe in $H$ on the basis of evidence).

@@

One widely known factor is the contribution of statistical methods to theoretical risk.
Theoretical risk $P(\neg{}E|\neg{}H)$ is deeply connected with statistical methods, because it is related to the type-I error rate in statistical hypothesis testing $P(E|\neg{}H)$ by $P(\neg{}E|\neg{}H) = 1 - P(E|\neg{}H)$, if you consider the overly simplistic case where the research hypothesis is equal to the statistical alternative-hypothesis because then the nill-hypothesis is $\neg{}H$.
Because many researchers are familiar with the type-I error rate, it can be helpful to remember this connection to theoretical risk.
Researchers who choose a smaller type-I error rate can be more sure of their results, if significant, because the theoretical risk is higher.
However, this connection should not be overinterpreted for two reasons.
First, according to most interpretations of null hypothesis testing, the absence of a significant result should not generally be interpreted as evidence against the hypothesis [@mayoStatisticalInferenceSevere2018, 5.3].
Second, the research hypothesis seldomly equals the statistical alternative-hypothesis (most research hypothesis are more specific then any value except zero).
In fact, it is entirely possible to assume the null hypothesis as a research hypothesis, as is commonly done in e.g., structural equation modelling, where the roles of detectability, theoretical risk and type-I/II error rate switch.
We argue that theoretical risk (and hence its complement, $P(E|\neg{}H)$) also encompasses factors outside the statistical realm, most notably the study design and broader analytical strategies.
Type-I error rate is the property of a statistical test under some assumptions, whereas theoretical risk is a researchers believe.
One may take such theoretical properties as a first starting point to form a substantive believe but surely researchers ought to take more factors into consideration.
For example, if a researcher believes that there might be confounding variables at play for the relation between two variables, that should decrease the theorethical risk; after all they might find an association purely on account of the confounders.

@@

The argument for a confirmatory research agenda is that by increasing theoretical risk we increase expected epistemic value, i.e., moving to the right on the x-axis in Figure \ref{fig:posterior} increases posterior probability (on the y-axis).
However, if a hypothesis in a certain study has low theoretical risk, there is not much researchers can do about it.
However, studies do not only differ by how high the theoretical risk is but also by how certain the recipient is about the theoretical risk.
A study that has a very high theoretical risk (e.g., `r (1-confirmatory_tau)*100`% chance that if the hypothesis is wrong, evidence in its favor will be observed,) but has also maximum uncertainty will result in a posterior probability of `r round(confirmatory_uncertain*100)`%, while the same study with maximum certainty will result in `r round(confirmatory_certain*100)`% posterior probability.
The other factors (detectability, prior beliefs, measure of epistemic value) and, therefore, the extent of the benefit varies, of course, with the specifics of the study.
Crucially, even studies with some exploratory aspects benefit from preregistration, e.g., in this scenario with a $\tau$ = `r some_exploration_tau` (false positive rate of `r 1-some_exploration_tau`) moving from uncertain to certain increases the posterior from `r round(some_exploration_uncertain, 2)` to `r round(some_exploration_certain, 2)`.

@@

In case 1, there is no theoretical risk ($P(\neg{}E|\neg{}H) = 0$).
If we know that no matter what, the results engineered to support the hypothesis, there is no reason to collect data.
Case 2 has a similar problem.
In fact, both cases should be problematic to anyone who subscribes to the statistical relevancy condition because if we choose the hypothesis in accordance with the data or vice versa, without restrictions, they are not related anymore (i.e., observing the data does not tell us anything about the hypothesis and the other way around).
Case 3, is different since here the theoretical risk is not necessarily low but simply uncertain and perhaps represented by the dotted line in Figure \ref{fig:posterior}/\ref{fig:other-measures}.
In case 4, the theoretical risk is neither zero (unless the researcher plans to do everything in their power to obtain favourable results, then we have a particular instance case of 1) nor high (that is what makes it exploratory).
However, we can use statistical properties, simulation or resampling methods, together with scientific reasoning, to get a reasonably certain evaluation of the theoretical risk and hence are in a somewhat favourable position (i.e., close to the solid line in n Figure \ref{fig:posterior}/\ref{fig:other-measures}).
This favorable position leads us to recommend preregistration of exploratory studies.
Case 5 shares the neither zero nor high theoretical risk of case 4 but has additional uncertainty about how much exploration was going on (how hard did researcher try to come up with favourable results).
Its low theoretical risk combined with high uncertainty about it, make it difficult to produce compelling evidence.

@@

With the goal to facilitate, rigorous exploration, we have proposed a workflow for preregistration called _preregistration as code_ (PAC) elsewhere [@peikertReproducibleResearchTutorial2021].
In a PAC, researchers use computer code for the planned analysis as well as a verbal description of theory and methods for the preregistration.
This combination is facilitated by dynamic document generation, where the results of the code, such as numbers, figures, and tables, are inserted automatically into the document.
The idea is that the preregistration already contains "mock results" based on simulated or pilot data, which are replaced after the actual study data becomes available.
Such an approach dissolves the distinction between the preregistration document and the final scientific report.
Instead of separate documents, preregistration, and final report are different versions of the same underlying dynamic document.
Deviations from the preregistration can therefore be clearly (and if necessary, automatically) isolated, highlighted, and inspected using version control.
Crucially, because the preregistration contains code, it may accommodate many different data patterns, i.e., it may be exploratory.
However, while a PAC does not limit the extent of exploration, it is very specific about the probability to generate evidence even when the theory does not hold (theoretical risk).
Please note that while PAC is ideally suited to reduce uncertainty about theoretical risk, other more traditional forms of preregistration are also able to advance this goal.

>>>>

We are interested in why preregistrations should change researchers' evaluation of evidence.
Applying a Bayesian framework allows us to investigate our research question most straightforwardly. 
Specifically, it allows us to model changes in subjective degrees of belief due to preregistration or, more simply, "persuasion".
Please note that our decision to adopt a Bayesian philosophy of science does not make assumptions about the statistical methods researchers use.
In fact, this conceptualization is intentionally as minimal as possible to be compatible with a wide range of philosophies of science and statistical methods researchers might subscribe to.
However, we should note that Popperians would be appalled that we are content with positive inductive inferences (but we dislike "failing to disprove" too much to give it up), and Neopopperians would flinch that we assign probabilities to beliefs (we are fond of calculating things).
While the latter move is not strictly necessary it allows us to connect the more abstract considerations more closely with the behavior of researchers.

@@

The posterior probability $P(H|E)$ is of great relevance since it is often used directly or indirectly as a measure of confirmation of a hypothesis.
In the tradition of Carnap, in its direct use, it is called confirmation as firmness; in its relation to the a priori probability $P(H)$, it is called _increase in firmness_ [@carnapLogicalFoundationsProbability1950, preface to the 1962 edition].
We concentrate on the posterior probability because of its simplicity but take it only as one example of a possible measure.
In reality, surely researchers differ in what function they apply to judge evidence and it is often most fruitful to compare more than two competing hypotheses.
The goal is therefore to reason about the space of possible measures researchers might apply.
However, since it is reasonable that any measure of confirmation increases monotonically with an increase in posterior probability $P(H|E)$, we might well take it to illustrate our reasoning.

In short, we want to increase posterior probability $P(H|E)$.
Increases in posterior probability $P(H|E)$ are associated with increased epistemic value, of which we want to maximize the expectation.
So how can we increase posterior probability?
The Bayes formula yields three components that influence confirmation, namely $P(H)$, $P(E|H)$ and $P(E)$.
The first option leads us to the unsurprising conclusion that higher a priori probability $P(H)$ leads to higher posterior probability $P(H|E)$.
If a hypothesis is more probable to begin with, observing evidence in its favor will result in a hypothesis that is more strongly confirmed, all else being equal.
However, the prior probability of a hypothesis is nothing our study design can change.
The second option is equally reasonable; that is, an increase in $P(E|H)$ leads to a higher posterior probability $P(H|E)$.
$P(E|H)$ is the probability of obtaining evidence for a hypothesis when it holds.
We call this probability of detecting evidence, given that the hypothesis holds "detectability."
Consequently, researchers should ensure that their study design allows them to find evidence for their hypothesis, in case it is true.
When applied strictly within the bounds of null hypothesis testing, detectability is equivalent to power (or the complement of type-II error rate).
However, while detectability is of great importance for study design, it is not directly relevant to what a preregistration is comunicating to other researchers.
However, we latter discuss how issues detectability must be considered for the issue of preregistration.
Thus, $P(E)$ remains to be considered.
Since $P(E)$ is the denominator, decreasing it can increase the posterior probability.
In other words, high risk, high reward.

@@

The last statement sounds like a truism but is directly related to Popper's seminal criterion of demarcation.
He stated that if it is impossible to prove that a hypothesis is false ($P(\neg{}E|\neg{}H) = 0$, theoretical risk is zero), it cannot be considered a scientific hypothesis [@popperLogicScientificDiscovery2002, p. 18].
We note these relations to underline that the Bayesian rationale we apply here is able to reconstruct many commonly held views on riskiness and epistemic value (but Popper would disagree with attaching any form of believe in $H$ on the basis of evidence beyond failing to disprove it).

@@

One widely known factor is the contribution of statistical methods to theoretical risk.
Theoretical risk $P(\neg{}E|\neg{}H)$ is deeply connected with statistical methods, because it is related to the type-I error rate in statistical hypothesis testing $P(E|\neg{}H)$ by $P(\neg{}E|\neg{}H) = 1 - P(E|\neg{}H)$, if you consider the overly simplistic case where the research hypothesis is equal to the statistical alternative-hypothesis because then the nill-hypothesis is $\neg{}H$.
Because many researchers are familiar with the type-I error rate, it can be helpful to remember this connection to theoretical risk.
Researchers who choose a smaller type-I error rate can be more sure of their results, if significant, because the theoretical risk is higher.
However, this connection should not be overinterpreted for two reasons.
First, according to most interpretations of null hypothesis testing, the absence of a significant result should not generally be interpreted as evidence against the hypothesis [@mayoStatisticalInferenceSevere2018, 5.3].
Second, the research hypothesis seldomly equals the statistical alternative-hypothesis (most research hypothesis are more specific then "any value except zero").
In fact, it is entirely possible to assume the null hypothesis as a research hypothesis, as is commonly done in e.g., structural equation modelling, where the roles of detectability, theoretical risk and type-I/II error rate switch.
We argue that theoretical risk (and hence its complement, $P(E|\neg{}H)$) also encompasses factors outside the statistical realm, most notably the study design and broader analytical strategies.
Type-I error rate is the property of a statistical test under some assumptions, whereas theoretical risk is a researchers believe.
One may take such theoretical properties as a first starting point to form a substantive believe but surely researchers ought to take more factors into consideration.
For example, if a researcher believes that there might be confounding variables at play for the relation between two variables, that should decrease the theorethical risk; after all they might find an association purely on account of the confounders.

@@

The argument for a confirmatory research agenda is that by increasing theoretical risk we increase expected epistemic value, i.e., moving to the right on the x-axis in Figure \ref{fig:posterior} increases posterior probability (on the y-axis).
However, if a hypothesis in a certain study has low theoretical risk, there is not much researchers can do about it.
However, studies do not only differ by how high the theoretical risk is but also by how certain the recipient is about the theoretical risk.
A study that has a very high theoretical risk (e.g., `r (1-confirmatory_tau)*100`% chance that if the hypothesis is wrong, evidence in its favor will be observed,) but has also maximum uncertainty will result in a posterior probability of `r round(confirmatory_uncertain*100)`%, while the same study with maximum certainty will result in `r round(confirmatory_certain*100)`% posterior probability.
The other factors (detectability, prior beliefs, measure of epistemic value) and, therefore, the extent of the benefit varies, of course, with the specifics of the study.
Crucially, even studies with some exploratory aspects benefit from preregistration, e.g., in this scenario with a $\tau$ = `r some_exploration_tau` (false positive rate of `r 1-some_exploration_tau`) moving from uncertain to certain increases the posterior from `r round(some_exploration_uncertain, 2)` to `r round(some_exploration_certain, 2)`.
We find it helpful to calculate some example because of the nonlinear nature of the evidence functions.

@@

In case 1, there is no theoretical risk ($P(\neg{}E|\neg{}H) = 0$).
If we know that no matter what, the results engineered to support the hypothesis, there is no reason to collect data.
Case 2 has a similar problem.
After all, the hypothesis that it had to happen the way it did happen is irrefutable.
In fact, both cases should be problematic to anyone who subscribes to the statistical relevancy condition because if we choose the hypothesis in accordance with the data or vice versa, without restrictions, they are not related anymore (i.e., observing the data does not tell us anything about the hypothesis and the other way around).
Case 3, is different since here the theoretical risk is not necessarily low but simply uncertain and perhaps represented by the dotted line in Figure \ref{fig:posterior}/\ref{fig:other-measures}.
In case 4, the theoretical risk is neither zero (unless the researcher plans to do everything in their power to obtain favourable results, then we have a particular instance case of 1) nor high (that is what makes it exploratory).
However, we can use statistical properties, simulation or resampling methods, together with scientific reasoning, to get a reasonably certain evaluation of the theoretical risk and hence are in a somewhat favourable position (i.e., close to the solid line in n Figure \ref{fig:posterior}/\ref{fig:other-measures}).
This favorable position leads us to recommend preregistration of exploratory studies.
Case 5 shares the neither zero nor high theoretical risk of case 4 but has additional uncertainty about how much exploration was going on (how hard did researcher try to come up with favourable results).
Its low theoretical risk combined with high uncertainty about it, make it difficult to produce compelling evidence.

@@

With the goal to facilitate rigorous exploration, we have proposed a workflow for preregistration called _preregistration as code_ (PAC) elsewhere [@peikertReproducibleResearchTutorial2021].
In a PAC, researchers use computer code for the planned analysis as well as a verbal description of theory and methods for the preregistration.
This combination is facilitated by dynamic document generation, where the results of the code, such as numbers, figures, and tables, are inserted automatically into the document.
The idea is that the preregistration already contains "mock results" based on simulated or pilot data, which are replaced after the actual study data becomes available.
Such an approach dissolves the distinction between the preregistration document and the final scientific report.
Instead of separate documents, preregistration, and final report are different versions of the same underlying dynamic document.
Deviations from the preregistration can therefore be clearly (and if necessary, automatically) isolated, highlighted, and inspected using version control.
Crucially, because the preregistration contains code, it may accommodate many different data patterns, i.e., it may be exploratory.
However, while a PAC does not limit the extent of exploration, it is very specific about the probability to generate evidence even when the theory does not hold (theoretical risk).
Please note that while PAC is ideally suited to reduce uncertainty about theoretical risk, other more traditional forms of preregistration are also able to advance this goal.
